---
title: "Homework Assignment #2"
subtitle: "Exploring patterns of environmental justice"
author: "Jordan Sibley"
date: last-modified
execute: 
  eval: false
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

## Set up 
```{r}
# load libraries 
library(tidyverse)
library(tmap)
library(dplyr)
library(sf)
```

```{r}
# read in data

```


## Part 1: Legacy of redlining in current environmental (in)justice

Your first task is to explore historical redlining in Los Angeles and its legacy on present-day environmental justice.

#### Description

1.  a map of historical redlining neighborhoods, including:

-   neighborhoods colored by HOLC grade
-   an appropriate base map

2.  a table summarizing the percent of current census block groups within each HOLC grade (or none)

3.  a set of figures summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables:

-   \% low income
-   percentile for Particulate Matter 2.5
-   percentile for low life expectancy

4.  brief paragraph reflecting on these results

## Part 2: Legacy of redlining in biodiversity observations

Your second task is to explore the legacy of historical redlining in Los Angeles on the collection of bird observations.

#### Description

For this assignment, you must produce the following based on observations from 2022:

1.  a figure summarizing the percent of observations within redlined neighborhoods within each HOLC grade
2.  a brief paragraph explaining whether these results match the findings from Ellis-Soto et al. 2023
